geochemistry

### 2025-02-05 geochemistry.qmd
pXRF: Improved tables and graphs; organised code and chunks.

library(readr)
library(dplyr)
library(janitor)
library(tidyr)
library(gt)
library(ggplot2)
library(cowplot)
# Load required libraries
library(tidyverse)
library(ggfortify)
library(FactoMineR)
library(factoextra)

@@

## DB setup
pxrf_raw <- pxrf_raw |> 
  filter(!Sample_ID %in% c("I19-1751", "H21-2126")) # Removing 2 altered samples
pxrf_analysis <- pxrf_raw |> 
  select("Sample_ID", "pXRF_ID", "Formation/Type",
         "MgO", "Al2O3", "SiO2", "P2O5", "S", "Cl", "K2O", "CaO", "TiO2", "Fe2O3", "Cu",
         "Sr", "Zr") |>  # Removed elements with 0 or <LOD values in all samples or in high percentages in all groups
  mutate(`Formation/Type` = case_when(
    `Formation/Type` == "TYPE 1" ~ "T1",
    `Formation/Type` == "TYPE 2" | `Formation/Type` == "TYPE 2B" | `Formation/Type` == "TYPE 2R" ~ "T2",
    `Formation/Type` == "TYPE 3" ~ "T3",
    `Formation/Type` == "TYPE 4" ~ "T4",
    `Formation/Type` == "TYPE 5" ~ "T5",
    `Formation/Type` == "TYPE 6" | `Formation/Type` == "TYPE 6LLB" | `Formation/Type` == "TYPE 6E" ~ "T6",
    `Formation/Type` == "TYPE 7" ~ "T7",
    `Formation/Type` == "TYPE 8" ~ "T8",
    `Formation/Type` == "TYPE 9" ~ "T9",
    `Formation/Type` == "TYPE 10" ~ "T10",
    `Formation/Type` == "TYPE 11" ~ "T11",
    TRUE ~ `Formation/Type`
  ))

@@

@@

@@

# Table to see the occurrence of 0 or <LOD values within each element and by rm type, to choose the elements in which to apply statistics.
# Combine the multiple columns into a single table
combined_table <- pxrf_raw |>
  select("Sample_ID", "pXRF_ID", "Formation/Type",
         "MgO", "Al2O3", "SiO2", "P2O5", "S", "Cl", "K2O", "CaO", "TiO2", "V", "MnO", "Fe2O3", "Cu", "As", "Rb",
         "Sr", "Y", "Zr", "Mo", "Ag", "Ba", "Ce", "W", "Au", "Pb") |> 
  mutate(across(
    .cols = c("MgO", "Al2O3", "SiO2", "P2O5", "S", "Cl", "K2O", "CaO", "TiO2", "V", "MnO", "Fe2O3", "Cu", "As", "Rb",
         "Sr", "Y", "Zr", "Mo", "Ag", "Ba", "Ce", "W", "Au", "Pb"),
    .fns = ~ case_when(
      . == 0 | . == "< LOD" ~ "No",
      . > 0 ~ "Yes",
      TRUE ~ as.character(.))
    )) |> 
  mutate(`Chert type` = case_when(
    `Formation/Type` == "TYPE 2B" | `Formation/Type` == "TYPE 2R" ~ "TYPE 2",
    `Formation/Type` == "TYPE 6LLB" | `Formation/Type` == "TYPE 6E" ~ "TYPE 6",
    TRUE ~ `Formation/Type`))

@@

pxrf_average <- pxrf_analysis |> 
  select(-Sample_ID, -pXRF_ID) |> 
  mutate(across(-`Formation/Type`, ~ as.numeric(.))) |>    # Convert valid numbers
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), 0, .))) |>  # Replace NAs with 0
  group_by(`Formation/Type`) |> 
  summarise(across(where(is.numeric), ~ round(mean(.), 4)))
  pivot_wider(names_from = `Formation/Type`,
              values_from = c("MgO", "Al2O3", "SiO2", "P2O5", "S", "Cl", "K2O", "CaO", "TiO2", "Fe2O3", "Cu",
         "Sr", "Zr"))

average_MgO <- pxrf_average |> 
  ggplot(aes(x = `Formation/Type`, y = MgO)) +
  geom_bar(stat = "identity") +
  theme_minimal() + 
  labs(x = " ",y = "MgO")

average_Al2O3 <- pxrf_average |> 
  ggplot(aes(x = `Formation/Type`, y = Al2O3)) +
  geom_bar(stat = "identity") +
  theme_minimal() + 
  labs(x = " ",y = "Al2O3")

average_SiO2 <- pxrf_average |> 
  ggplot(aes(x = `Formation/Type`, y = SiO2)) +
  geom_bar(stat = "identity") +
  theme_minimal() + 
  labs(x = " ",y = "SiO2")

average_P2O5 <- pxrf_average |> 
  ggplot(aes(x = `Formation/Type`, y = P2O5)) +
  geom_bar(stat = "identity") +
  theme_minimal() + 
  labs(x = " ",y = "P2O5")

average_S <- pxrf_average |> 
  ggplot(aes(x = `Formation/Type`, y = S)) +
  geom_bar(stat = "identity") +
  theme_minimal() + 
  labs(x = " ",y = "S")

average_Cl <- pxrf_average |> 
  ggplot(aes(x = `Formation/Type`, y = Cl)) +
  geom_bar(stat = "identity") +
  theme_minimal() + 
  labs(x = " ",y = "Cl")

average_K2O <- pxrf_average |> 
  ggplot(aes(x = `Formation/Type`, y = K2O)) +
  geom_bar(stat = "identity") +
  theme_minimal() + 
  labs(x = " ",y = "K2O")

average_CaO <- pxrf_average |> 
  ggplot(aes(x = `Formation/Type`, y = CaO)) +
  geom_bar(stat = "identity") +
  theme_minimal() + 
  labs(x = " ",y = "CaO")

average_TiO2 <- pxrf_average |> 
  ggplot(aes(x = `Formation/Type`, y = TiO2)) +
  geom_bar(stat = "identity") +
  theme_minimal() + 
  labs(x = " ",y = "TiO2")

average_Fe2O3 <- pxrf_average |> 
  ggplot(aes(x = `Formation/Type`, y = Fe2O3)) +
  geom_bar(stat = "identity") +
  theme_minimal() + 
  labs(x = " ",y = "Fe2O3")

average_Sr <- pxrf_average |> 
  ggplot(aes(x = `Formation/Type`, y = Sr)) +
  geom_bar(stat = "identity") +
  theme_minimal() + 
  labs(x = " ",y = "Sr")

>>>>

library(readr)
library(dplyr)
library(janitor)
library(tidyr)
library(gt)
library(ggplot2)
library(cowplot)
library(tidyverse)
library(ggfortify)
library(FactoMineR)
library(factoextra)

@@

## DB setup
pxrf_raw <- pxrf_raw |> 
  filter(!Sample_ID %in% c("I19-1751", "H21-2126")) # Removing 2 altered samples
pxrf_analysis <- pxrf_raw |> 
  select("Sample_ID", "pXRF_ID", "Formation/Type",
         "Al2O3", "SiO2", "P2O5", "S", "Cl", "K2O", "CaO", "TiO2", "Fe2O3", "Cu",
         "Sr", "Zr") |>  # Removed elements with 0 or <LOD values in all samples or in high percentages in all groups
  mutate(`Formation/Type` = case_when(
    `Formation/Type` == "TYPE 1" ~ "T1",
    `Formation/Type` == "TYPE 2" | `Formation/Type` == "TYPE 2B" | `Formation/Type` == "TYPE 2R" ~ "T2",
    `Formation/Type` == "TYPE 3" ~ "T3",
    `Formation/Type` == "TYPE 4" ~ "T4",
    `Formation/Type` == "TYPE 5" ~ "T5",
    `Formation/Type` == "TYPE 6" | `Formation/Type` == "TYPE 6LLB" | `Formation/Type` == "TYPE 6E" ~ "T6",
    `Formation/Type` == "TYPE 7" ~ "T7",
    `Formation/Type` == "TYPE 8" ~ "T8",
    `Formation/Type` == "TYPE 9" ~ "T9",
    `Formation/Type` == "TYPE 10" ~ "T10",
    `Formation/Type` == "TYPE 11" ~ "T11",
    TRUE ~ `Formation/Type`
  ))

@@

```{r}
#| label: fig-lod
#| fig-cap: A barplot.

# Table to see the occurrence of 0 or <LOD values within each element and by rm type, to choose the elements in which to apply statistics.
# Combine the multiple columns into a single table
lod_tab <- pxrf_raw |>
  select(-contains("Err")) |>
  select(-pXRF_ID, -Application, -Method, -ElapsedTime, -`Cal Check`) |>
  mutate(across(.cols = MgO:U,
                .fns = ~ case_when(
                  . == 0 ~"Success 0",
                  . > 0 ~ "Success",
                  TRUE ~ as.character(.))
                ))
# Reshape the data into a long format
lod_tab_long <- lod_tab |>
  pivot_longer(cols = MgO:U,  # Assuming these columns contain the 'Success' and 'LOD' values
               names_to = "Element",  # New column for the names of elements
               values_to = "Value")   # New column for the corresponding Success/LOD values

lod_tab_summary <- lod_tab_long |>
  group_by(Element, Value) |>
  summarise(Count = n(), .groups = "drop")  # Count the occurrences of Success and LOD for each Element

# Now, reshape the data into a wide format using pivot_wider
lod_tab_table <- lod_tab_summary |>
  pivot_wider(names_from = Value, values_from = Count, values_fill = list(Count = 0))  # Fill missing values with 0

lod_tab_summary <- lod_tab_long |>
  group_by(Element, Value) |>
  summarise(Count = n(), .groups = "drop")  # Count occurrences of each Value per Element

lod_tab_summary <- lod_tab_summary |>
  count(Element, Value)

# Create the bar plot
ggplot(lod_tab_summary, aes(x = Element, y = Count, fill = Value)) +
  geom_bar(stat = "identity", position = "stack") +  # Use stat = "identity" to use summarized counts
  scale_fill_manual(values = c("Success" = "darkseagreen", "Success 0" = "darkseagreen2", "< LOD" = "indianred3")) +
  labs(title = "Summarized Success and LOD Counts by Element",
       y = "Count", 
       fill = " ") +
  theme_minimal() +
 # facet_wrap(~ Element, scales = "free_y") +  # Facet by Element with independent y-axes
  theme(axis.text.x = element_text(hjust = 1),
        axis.title.x = element_blank()) +
  #geom_text(data = lod_tab_summary_filtered, aes(label = Count), position=position_stack(vjust=0.5), colour="white", size = 3) +
  coord_flip() +
  scale_y_continuous(breaks = seq(0, max(lod_tab_summary$Count), by = 20))

@@

pxrf_average <- pxrf_analysis |> 
  select(-Sample_ID, -pXRF_ID) |>
  mutate(`Formation/Type` = case_when(
    `Formation/Type` == "T2" | `Formation/Type` == "T3" | `Formation/Type` == "T4" | `Formation/Type` == "T5" ~ "T2-5", TRUE ~ `Formation/Type`
  )) |> 
  mutate(`Formation/Type` = factor(`Formation/Type`, levels = c("TW", "LJW", "MJE", "UJW", "UJE", "CPT", "UCA", "T1", "T2-5", "T6", "T7", "T8", "T9", "T10", "T11"))) |>
  mutate(across(-`Formation/Type`, ~ as.numeric(.))) |>    # Convert valid numbers
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), 0, .))) |>  # Replace NAs with 0
  group_by(`Formation/Type`) |> 
  summarise(across(where(is.numeric),
                   .fns = list(mean = ~ round(mean(.x, na.rm = TRUE), 4), 
                               sd = ~ round(sd(.x, na.rm = TRUE), 4)), 
                   .names = "{.col}_{.fn}"))

@@

combined_table <- pxrf_raw |>
  select("Sample_ID", "pXRF_ID", "Formation/Type",
         "MgO", "Al2O3", "SiO2", "P2O5", "S", "Cl", "K2O", "CaO", "TiO2", "V", "MnO", "Fe2O3", "Cu", "As", "Rb",
         "Sr", "Y", "Zr", "Mo", "Ag", "Ba", "Ce", "W", "Au", "Pb") |> 
  mutate(across(
    .cols = c("MgO", "Al2O3", "SiO2", "P2O5", "S", "Cl", "K2O", "CaO", "TiO2", "V", "MnO", "Fe2O3", "Cu", "As", "Rb",
         "Sr", "Y", "Zr", "Mo", "Ag", "Ba", "Ce", "W", "Au", "Pb"),
    .fns = ~ case_when(
      . == 0 | . == "< LOD" ~ "No",
      . > 0 ~ "Yes",
      TRUE ~ as.character(.))
    )) |> 
  mutate(`Chert type` = case_when(
    `Formation/Type` == "TYPE 2B" | `Formation/Type` == "TYPE 2R" ~ "TYPE 2",
    `Formation/Type` == "TYPE 6LLB" | `Formation/Type` == "TYPE 6E" ~ "TYPE 6",
    TRUE ~ `Formation/Type`))

@@
